### 2018-01-15 paper/response_to_reviewers.tex

\item{\textcolor{blue}{Another concern I have is about how this R package sits within the current analytical toolkit available.
How does this new package relate to existing packages available for R?
Claddis and geomorph are mentioned and discussed because they deal with bespoke data formats that need to be dealt with prior to analysis.
But are there others?
How does the package relate to the ‘vegan’ R package for multivariate analysis of ecological communities?
Within there are metrics for community dissimilarity, metrics from ordinations etc.
Vegan is a popular package for analysis of multivariate datasets and seems to be relevant here with similar functionalities but not discussed.
Similarly, the R package ‘diverse’ has a function called disparity, and the ‘geiger’ function disparity implements the disparity-through-time. }}
\label{add_geiger_vegan}

I have added two wrapper functions to use the popular functions \texttt{vegan::adonis} and \texttt{geiger::dtt} to the package (\texttt{dispRity.adonis} and \texttt{dispRity.dtt}).
Also, I have generalised the distances calculations using the \texttt{vegan::vegdist} function that allows for more distances than \texttt{stats::dist} to be calculated.

I've updated the manual and the example to describe these new functions:

\textit{Similarly to the example above, it is also possible to statistically test this hypothesis using, for example, permutational multivariate ANOVA (PERMANOVA; Anderson, 2001b) through the adonis.disprity function that is a wrapper of the vegan::adonis function (Oksanen et al., 2007) for dispRity objects [TABLE 4]. To answer our specific question above: yes there is an effect of time on morphological disparity (an increase) in this dataset. Note that in this case, the function outputs different warnings on the usage of such test and the eventual data not used in the test. Additionally, the test is not applied to the bootstrapped data and thus might be sensitive to outliers and sampling size.} lines 333-342.

@@

I have however modified the available models to better describe this limitation and I have added two new models that would approximate ancestral reconstruction along edges.
I have modified the name of the two previous ``punctuated'' and ``gradual'' models to ``random'' and ``proximity'' to reflect more accurately how they select either the ACCTRAN or DELTRAN information.
I've added two more accurate models (now named ``equal.split'' and ``gradual.split'') to reflect the changes along the branches.
These two algorithm now allows to use both the ACCTRAN and DELTRAN information with a linked probability to use either of them per bootstrap draw.
For example, the ``equal.split'' model will have a 0.5 probability of using either node information for each bootstrap draw; the ``gradual.split'' model transforms the branch length were the slice occurs into a probability of drawing either of the nodes information (e.g. if the slices occurs 3/4 along the branch from A to B, the model will draw the information from A and B with a probability of respectively 0.25 and 0.75).
For example, the median disparity over 1000 bootstrap pseudo-replication along the edge (AB) will integrate 0.75 information from A and 0.25 from B thus approximating the ``intermediate'' between A and B (when averaged ).

I've added these models descriptions in the manual and in the manuscript:

\begin{itemize}
    \item \textit{``random'' where the data chosen along the branch is randomly chosen between the descendant or the ancestor}
    \item \textit{``proximity'' where the data chosen along the branch is either the descendant or the ancestor depending on branch length}
    \item \textit{``equal.split'' where the data chosen along the branch is both the descendant and the ancestor with an even probability}
    \item \textit{``gradual.split'' where the data chosen along the branch is both the descendant and the ancestor with a probability depending on branch length}
\end{itemize}

\textit{Note that the four first models are a proxy for punctuated evolution and the two last ones for gradual evolution.
There is a trade-off between precision and accuracy when using the time-slicing method: a higher number of slices increases the precision of the disparity analysis but also decreases accuracy.
See Guillerme \& Cooper \textit{in review} for full description.} lines 167-176.

@@

I've changed ``subsample'' to ``subset'' throughout the package, the manual and the paper.

I also agree with this comment: this method is a trade-off between the number of slices.
(1) too few slices decrease resolution and veer towards the time binning (which could increase "significance" by comparing too different time subsamples);
(2) too many slices will increase the resolution be will create much more pseudo replication and thus increase type I error.

I've added a warning in the manual and the manuscript:

\textit{There is a trade-off between precision and accuracy when using the time-slicing method: a higher number of slices increases the precision of the disparity analysis but also decreases accuracy.} line 173-175.

@@

\textit{Note that the input matrix is not restricted to an ordinated matrix but can be any kind of matrix as long as its rows represent elements (e.g. the multidimensional space can be a distance matrix; Close et al., 2015).} lines 109-111.

\item{\textcolor{blue}{Picking up on the first point from above, there are also additional post-ordination disparity metrics that have been used in the literature but are not available here. For example, hypercube volume, minimum spanning tree length etc. I would also urge the author to cite the work of Mike Foote, who made some early important contributions to disparity in general, and has written papers that discuss some of these potential metrics (e.g., Foote 1997, Annual Review of Ecology and Systematics, 28, 129–). I believe Wills (2001, cited) also discusses a long list of metrics.}}

I agree, there are many more metrics commonly used for disparity analysis not implemented in this package.
This was a conscious choice: due to the really high number of metrics available (especially when combining different ones), I designed the package to be modular hence allowing users to implement their own metrics (see response to reviewer 1 minor comment \ref{table_missing}).
I have add the hypercube volume description as an example of user defined metric in the manuscript:

\textit{When multiple functions are passed to the} \texttt{metric} \textit{argument, they are sorted by dimension-level and applied in decreasing order to the data.
For example, if the metric is defined as} \texttt{metric = c(prod, ranges)} \textit{(the hypercube volume), the} \texttt{ranges} \textit{function (dimension-level 2) is first applied to data (the range of each dimensions) and the function} \texttt{prod} \textit{is then applied to the results (}\texttt{prod(ranges(data))}\textit{).
One can also directly pass a function description to the} \texttt{metric}\textit{ argument (e.g.} \texttt{metric = function(x) mean(dist(x)$\hat{\mkern6mu}$2)} \textit{ for the average squared pairwise distance).} lines 127-134.

@@

\label{ANOVA}

I agree that this ANOVA is a garbage-in-garbage-out test and used it more for illustrating the package's possibilities rather than show-casing disparity analysis best (worst?) practices.
Following this reviewer's comments with which I totally agree, I have completely removed the ANOVA example (from the paper, the manuals and the vignette), replacing it in the paper with a PERMANOVA (based on \texttt{vegan::adonis}) which is more conventional.
As this reviewer points out, the ANOVA is indeed performed on the bootstrapped data rather than the observed data which, of course, artificially (and wrongly) increases the statistical power.
Although the ANOVA is not mentioned any more, I have however left the ability for \texttt{test.dispRity} to deal with \texttt{lm} type tests (with a warning being triggered).

I have implemented the reviewers' suggestion similary to the ``percentile bootstrap'' idea, allowing \texttt{test.dispRity} to work on either the pooled bootstrap distribution through the \texttt{concatenate = TRUE} argument (this was indeed erroneous in the previous ANOVA example but could be justifiable in other cases; e.g. when comparing distribution overlap probabilities) or by applying the test for each bootstrap pseudo-replicate with \texttt{concatenate = FALSE}. I've clarified this in the manuscript:

\textit{Additionally, the test is here performed on the pooled bootstrapped pseudo-replications which can increase the type I error.
It is possible to compare each bootstrap in a pairwise way without pooling the data by using the} \texttt{concatenate = FALSE} \textit{argument.
The results will then be a distribution of statistic and p values.} lines 298-302.

@@

I have changed the \texttt{test.dispRity} to now output the name of the statistic and the parameter in the output list (e.g. \texttt{statistic: t} and \texttt{parameter: df} for the example provided by the reviewer).
I have also highlighted in the manuscript one of the features of the function allowing to output the results in a format more familiar to the user using the option \texttt{details = TRUE} (returning the raw test results):

\textit{Note that by default the function only outputs the test's statistic, parameter (if parametric) and the p value. However, the raw test results can also be output using the option} \texttt{details = TRUE} \textit{in the function above.} lines 296-298.

@@

\begin{enumerate}

\item{\textcolor{blue}{I have several concerns regarding the content in the manuscript, however, primarily how it is pitched.
It seems unclear, to me at least, who the intended audience is?
At various points I get the impression that the author intends for the package to be useful, and therefore, presents it, to readers beyond those studying morphometry etc.
There are frequent references to ecology and related disciplines where similar approaches are used.
Yet, those in these broader fields will struggle to follow what is presented as the manuscript exclusively uses "morphospaces" and related terminology that will only be relevant to a specific subset of readers.If you want to aim for a more general audience, consider rewriting the manuscript in more general terminology; samples in rows, variables in columns, for example.
Perhaps include a table to relate terminology used in morphometry with that more commonly encountered in community ecology.
Consider adding an ecological example and removing, if length is an issue, content on the manual and vignette --- we don't really need a paragraph on that, just a sentence to highlight it exists.}}

Due to space constraint for this format of MEE manuscript, I have left the references to ecology in the introduction of the manuscript but removed any further references.
However, I have expanded the ecological example in the vignette which now include a test similar to the one in Diaz \textit{et al} 2016 (Science).
I therefore hope that the mention of ecology in the introduction will encourage ecology-focused researcher to explore the manual.

\textit{Note that more examples are available in the package manual (...).} lines 229-231

@@

I went with Legendre \& Legendre 1998's definition of ordinations (pp. 388) in the three major distinct categories: (1) principal components analysis (PCA), (2), principal coordinates and metric scaling (PCO, PCoA) and non metric dimensional scaling (NMDS, MDS), ignoring correspondence and factor analysis for clarity. See reviewer 2, minor suggestion \ref{PCO_cites}.

\item{\textcolor{blue}{L65--64 here are additional areas where referencing is poor. Consider Marti Anderson's work on multivariate ANOVA \& dispersions as earlier work, for example:
\begin{itemize}
\item Anderson, M. J. A new method for non-parametric multivariate analysis of variance. Austral Ecol. 26, 32–46 (2001)
\item Anderson, M. J. Distance-based tests for homogeneity of multivariate dispersions. Biometrics 62, 245–253 (2006).
\item Anderson, M. J., Ellingsen, K. E. \& McArdle, B. H. Multivariate dispersion as a measure of beta diversity. Ecol. Lett. 9, 683–693 (2006).
\end{itemize}
}}
\textcolor{blue}{For permutation tests, significant work has been done in this area by Cajo ter Braak, principally in his Canoco software and related publications.
Also comparing overlap in CIs is a terrible way to perform inference; for 95\% confidence intervals, if the intervals for two estimated means are as close as possible but not overlapping, the p value of the difference means test is 0.005, not 0.05. To achieve 0.05 you can allow considerable overlap, but how much is tricky to discern by eye.
As such you are at risk of making poor choices on the basis of perceived significance.}

I have changed this sentence by citing Anderson 2001 \textit{Austral Ecol.} and Manly 1997 \textit{Randomization, bootstrap and Monte Carlo methods in biology}.
I also totally agree with the poor choice of simply comparing CIs; unfortunately this method is not that uncommon common in palaeobiology.
I've removed the citation (to avoid pointing-and-shaming) and added a caveat to this way of comparing disparity:

\textit{[...] an equal variety of statistical tests such as non-parametric multivariate analyses of variance (NPMANOVA, Anderson 2001 ; e.g. in Brusatte et al. 2008) multidimensional permutation tests (Manly 1997 ; e.g. in Díaz et al. 2016) or even, less rigorously, by looking at the confidence interval overlaps between disparity measurements.} lines 69-73

@@

\item{\textcolor{blue}{L326 Where do the 300 samples come from as implied by the residual degrees of freedom. Table 3 suggests 15 + 9 + 13 = 37 samples in this analysis. I'm clearly missing something here. Also, this is an omnibus test, which if we believe the printed values (see previous re DF) only says there is some effect of `subsamples`. Can this analysis be extended to point estimates for mean variance (again assuming you can justify this sum of individual variances method) and post-hoc comparisons of pairwise `subsample` differences, which would be quite useful?}}

The high value of the samples, residuals and degrees of freedom comes from the test being performed on the pooled bootstrap values rather than the raw observations.
I have now removed any mentions to this dodge anova and replaced it with a more traditional PERMANOVA instead (see reviewer 1 major comment \ref{add_geiger_vegan} and reviewer 3, major comment \ref{ANOVA}).

\item{\textcolor{blue}{If you need to save space to add an ecological example, cut the *Manuals and vignettes* section and also the *Data simulations* section. The later is somewhat tangential to the main thrust of the paper and you cover this in the vignette/bookdown resources.}}
\label{remove_simulations}

I have removed the simulations part as suggested by this reviewer.
However, following the encouraging suggestions of reviewer 2 (major suggestion \ref{simulations}).
I have developed in much more details the simulation algorithms in the package manual.

### 2018-01-15 paper/dispRity_paper.tex
paper up

\noindent {\normalsize \sc Thomas Guillerme$^1$$^,$$^*$}\\
\noindent {\small \it 
$^1$Imperial College London, Silwood Park Campus, Department of Life Sciences, Buckhurst Road, Ascot SL5 7PY, United Kingdom.\\}
\end{center}
\medskip
\noindent{*\bf Corresponding author.} \textit{guillert@tcd.ie}\\  
\vspace{1in}

@@

\begin{table}
\resizebox{\textwidth}{!}
{%
    \begin{tabular}{p{2cm}|p{5cm}|p{1cm}|p{5cm}|p{5cm}}
        name & description & dim & definition & source \\
        \hline
        \texttt{ancestral. dist} & The distance between an element and its ancestor & 2 & $\sqrt{\sum_{i=1}^{n}{({k}_{n}-Ancestor_{n})^2}}$ & This package \\
        \texttt{centroids} & The distance between each element and a fixed point$^1$ of the space & 2 & $\sqrt{\sum_{i=1}^{n}{({k}_{n}-Centroid_{k})^2}}$ & This package \\
        \texttt{convhull. surface} & The surface of the convex hull & 1 & NA & \texttt{geometry::convhulln} \citep{barber1996quickhull,geometryR}\\
        \texttt{convhull. volume} & The volume of the convex hull & 1 & NA & \texttt{geometry::convhulln} \citep{barber1996quickhull,geometryR}\\
        \texttt{diagonal} & The greatest Euclidean distance & 1 & $\sqrt{\sum_{i=1}^{k}|max(k_i) - min(k_i)|}$ & This package \\
        \texttt{ellipse. volume}$^2$ & The volume of the ellipsoid & 1 & $\frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} (\lambda_{i}^{0.5})$ & This package; based on \cite{DonohueDim}\\
        \texttt{mode.val} & The modal value & 1 & NA & This package\\
        \texttt{n.ball. volume} & The hyper-spherical ($n$-ball) volume & 1 & $\frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} R$ & This package\\
        \texttt{pairwise. dist} & The pairwise distances between elements & 2 & NA & \texttt{vegan::vegdist} \citep{oksanen2007vegan}\\        
        \texttt{radius} & The radius of each dimensions & 2 & $|\frac{\sum_{i=1}^{n}k_i}{n} - f(\mathbf{v}k)|$ & This package\\
        \texttt{ranges} & The absolute ranges of each dimension & 2 & $|max(k_i) - min(k_i)|$ & This package \\
        \texttt{span.tree. length} & The minimal spanning tree length & 1 & $\sum(\mathrm{branch\ length})$ & \texttt{vegan::spantree} \citep{oksanen2007vegan}\\
        \texttt{variances} & The variance of each dimension & 2 & $\sigma^{2}{k_i}$ & This package \\
    \end{tabular}
}%
    % \caption{\small{Where $k$ is the number of dimensions, $n$ the number of elements, $\Gamma$ is the Gamma distribution, $\lambda_i$ is the eigenvalue of each dimensions, $\sigma^{2}$ is their variance and $Centroid_{k}$ is their mean, $Ancestor_{n}$ is the coordinates of the ancestor of element $n$, $f(\mathbf{v}k)$ is function to select one value from the vector $\mathbf{v}$ of the dimension $k$ (e.g. it's maximum, minimum, mean, etc.), $R$ is the radius of the sphere or the product of the radii of each dimensions ($\displaystyle\prod_{i=1}^{k}$ - for a hyper-ellipsoid). $^1$ by default that point is the centroid of the elements; $^2$ this function uses a fast estimation of the eigenvalue that only works in an ordinated space based on MDS or PCO/PCoA (\textit{not} PCA).}}
    % \label{Tab:metrics}
\end{table}

@@

%TODO: Add the proper sources for the convhulln and vegan packages

@@

The time-binning method groups elements by specific age range.
The time-slicing method works by using a phylogeny and looking at which taxa are present at any specific point in time.
This method thus requires the nodes to be part of the space, a dated phylogeny (chronogram) and which model to use when slicing through branches rather than tips and nodes.
When a slice occurs not on a tip or a node, six methods are available to select either the descendent or the ancestor's node/tip as an element for this time slice:
\begin{itemize}
    \item ``acctran'' where the data chosen along the branch is always the one of the descendant
    \item ``deltran'' where the data chosen along the branch is always the one of the ancestor
    \item ``random'' where the data chosen along the branch is randomly chosen between the descendant or the ancestor
    \item ``proximity'' where the data chosen along the branch is either the descendant or the ancestor depending on branch length
    \item ``equal.split'' where the data chosen along the branch is both the descendant and the ancestor with an even probability
    \item ``gradual.split'' where the data chosen along the branch is both the descendant and the ancestor with a probability depending on branch length
\end{itemize}

Note that the four first models are a proxy for punctuated evolution: the selected data is always either the one of the descendant or the ancestor.
The two last models are a proxy for gradual evolution: the data from both the descendant and the ancestor is used with an associate probability.
There is a trade-off between precision and accuracy when using the time-slicing method: a higher number of slices increases the precision of the disparity analysis but also increases the type I error (thus decreasing accuracy).
See Guillerme \& Cooper \textit{in review} for full description.

\subsection{Bootstrapping and rarefying}
Disparity measurement can be influenced by sampling \citep{Butler2012}.
To take this source of bias into account, it is common practice to bootstrap the multidimensional space (to minimise the effect of outliers) or/and to rarefy the data (to minimise the effect of different sample sizes in different subsets).
Additionally, if disparity is defined as a dimension-level 1 metric, it can be useful to measure it on bootstrapped data to obtain a distribution on which to perform statistical analyses.

Bootstrapping can be achieved by using the \texttt{boot.matrix} function which pseudo-replicates the space following two algorithms (1) the ``full'' algorithm where the bootstrapping is entirely stochastic ($n$ elements are replaced by any $m$ elements drawn from the data) and (2) the ``single'' algorithm where only one random element is replaced for each pseudo-replicate (somewhat akin to jackknife resampling).

Similarly, rarefaction can be achieved through the same \texttt{boot.matrix} function.
Rarefaction can be useful when looking at the effect of reducing the number of elements on the results of an analysis or for allowing the comparisons of subsets with the same number of elements in each.
In practice, rarefaction limits the number of elements to be drawn for each bootstrap replication: only $n-x$ elements are selected at each bootstrap replicate (where $x$ is the number of non-sampled elements).

\subsection{Interpreting results}
The functions above all generate a \disp object that can be summarised or plotted using the \texttt{S3} method functions \texttt{summary.dispRity} and \texttt{plot.dispRity} (provided disparity was calculated).
These results can also be analysed using the \texttt{test.dispRity} function for comparing subsets or testing hypotheses.

\subsubsection{Summarising and plotting}
The \texttt{summary.dispRity} and \texttt{plot.dispRity} functions allow users to set which central tendency and which quantiles should be represented.
Additionally, the \texttt{plot.dispRity} function graphically represents the summarised results using different representations: (1) ``continuous'' for displaying continuous disparity curves (in the case of disparity-through-time analysis for example) and (2) ``box'', ``lines'', or ``polygons'' to display them using boxplots, confidence interval lines or polygons respectively.
Additional arguments specific to \disp objects can also be used such as \texttt{observed} to display the observed disparity (i.e. non-bootstrapped) or \texttt{rarefaction} to only plot the disparity for a certain number of elements (i.e. the rarefaction level).
The function can also take any additional graphic arguments (\texttt{main}, \texttt{xlab}, \texttt{col}, etc...) from base \texttt{R}.

\subsubsection{Testing hypotheses}
The \texttt{test.dispRity} function allows users to test hypotheses on the disparity data (e.g. by comparing the different subsets to each other).
Similarly to the \disp function described above, this function can take any test defined by the user or from other R packages.
The \texttt{comparison} arguments indicates in which order (if any) the tests should be applied to the subsets: (1) ``pairwise'' for pairwise comparisons; (2)``referential'' for comparing the first subset to all the others; (3) ``sequential'' for comparing subsets sequentially (e.g. first against second, second against third, etc.); (4) `all'' for comparing all the subsets simultaneously (i.e. \texttt{disparity $\mathtt{\sim}$ subsets}) or (5) any list of pairs (or more) of subsets to compare.

@@

\section{Examples}
Multivariate analysis can be really useful for looking at multiple aspects of organisms diversity together.
For example, one classic univariate trait used in ecology and evolution can be the number of species present in a ecosystem or/and through time.
This trait can be useful to test whether some ecosystems or/and time periods have more species diversity than others.
In this case one can then test some hypothesis such as: environments with longer altitudinal gradients contain more species.

Conversely, in the case of species diversity, one can also look at another independent multivariate trait: the diversity of morphologies \citep[or disparity;][]{foote1991morphological}.
Using disparity, it is then also possible to assess whether one ecosystem or/and time period display more morphological variation.
Following the previous example, if disparity is also higher with longer altitudinal gradients, one could argue that the species richness is caused by an increased number of morphologies adapted to more ecological niches (or something akin).

The following example is based on a classical morphological disparity analysis.
Note that more examples are available in the package manual (\url{https://rawgit.com/TGuillerme/dispRity/master/inst/gitbook/_book/index.html}).

@@

In this example, the space is defined as a morphospace: the ordination of the distances among discrete morphological characters for 50 mammal species \citep[from][]{beckancient2014}.
Additionally, we can define disparity as the sum of the variances on each dimension \citep{foote1991morphological,Wills1994} that will represent an aspect of the the volume of the morphospace (i.e. a large sum of variances approximates a large morphospace and vice versa).

\subsection{Example}

The following example is based on the discrete morphological data described above. More examples, including based on other types of data are available in the package vignette.

@@

\section{Funding}
This work was funded by the European Research Council under the European Union's Seventh Framework Programme (FP/2007 - 2013)/ERC Grant Agreement number 311092 and the Discovery Project Grant number DP170103227.

>>>>

\noindent {\normalsize \sc Thomas Guillerme$^12$$^,$$^*$}\\
\noindent {\small \it 
$^1$Imperial College London, Silwood Park Campus, Department of Life Sciences, Buckhurst Road, Ascot SL5 7PY, United Kingdom.\
$^1$School of Biological Sciences, University of Queensland, St. Lucia, Australia.\\}

\end{center}
\medskip
\noindent{*\bf Corresponding author.} \textit{guillert@tcd.ie}\\  
\vspace{1in}

@@

\begin{table}
\resizebox{\textwidth}{!}
{%
    \begin{tabular}{p{2cm}|p{5cm}|p{1cm}|p{5cm}|p{5cm}}
        name & description & dim & definition & source \\
        \hline
        \texttt{ancestral. dist} & The distance between an element and its ancestor & 2 & $\sqrt{\sum_{i=1}^{n}{({k}_{n}-Ancestor_{n})^2}}$ & This package \\
        \texttt{centroids} & The distance between each element and a fixed point$^1$ of the space & 2 & $\sqrt{\sum_{i=1}^{n}{({k}_{n}-Centroid_{k})^2}}$ & This package \\
        \texttt{convhull. surface} & The surface of the convex hull & 1 & NA & \texttt{geometry::convhulln} \citep{barber1996quickhull,geometryR}\\
        \texttt{convhull. volume} & The volume of the convex hull & 1 & NA & \texttt{geometry::convhulln} \citep{barber1996quickhull,geometryR}\\
        \texttt{diagonal} & The greatest Euclidean distance & 1 & $\sqrt{\sum_{i=1}^{k}|max(k_i) - min(k_i)|}$ & This package \\
        \texttt{ellipse. volume}$^2$ & The volume of the ellipsoid & 1 & $\frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} (\lambda_{i}^{0.5})$ & This package; based on \cite{DonohueDim}\\
        \texttt{mode.val} & The modal value & 1 & NA & This package\\
        \texttt{n.ball. volume} & The hyper-spherical ($n$-ball) volume & 1 & $\frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} R$ & This package\\
        \texttt{pairwise. dist} & The pairwise distances between elements & 2 & NA & \texttt{vegan::vegdist} \citep{oksanen2007vegan}\\        
        \texttt{radius} & The radius of each dimensions & 2 & $|\frac{\sum_{i=1}^{n}k_i}{n} - f(\mathbf{v}k)|$ & This package\\
        \texttt{ranges} & The absolute ranges of each dimension & 2 & $|max(k_i) - min(k_i)|$ & This package \\
        \texttt{span.tree. length} & The minimal spanning tree length & 1 & $\sum(\mathrm{branch\ length})$ & \texttt{vegan::spantree} \citep{oksanen2007vegan}\\
        \texttt{variances} & The variance of each dimension & 2 & $\sigma^{2}{k_i}$ & This package \\
    \end{tabular}
}%
    \caption{\small{Where $k$ is the number of dimensions, $n$ the number of elements, $\Gamma$ is the Gamma distribution, $\lambda_i$ is the eigenvalue of each dimensions, $\sigma^{2}$ is their variance and $Centroid_{k}$ is their mean, $Ancestor_{n}$ is the coordinates of the ancestor of element $n$, $f(\mathbf{v}k)$ is function to select one value from the vector $\mathbf{v}$ of the dimension $k$ (e.g. it's maximum, minimum, mean, etc.), $R$ is the radius of the sphere or the product of the radii of each dimensions ($\displaystyle\prod_{i=1}^{k}$ - for a hyper-ellipsoid). $^1$ by default that point is the centroid of the elements; $^2$ this function uses a fast estimation of the eigenvalue that only works in an ordinated space based on MDS or PCO/PCoA (\textit{not} PCA).}}
    \label{Tab:metrics}
\end{table}

@@

@@

The time-binning method groups elements by specific age range.
The time-slicing method works by using a phylogeny and looking at which taxa are present at any specific point in time.
This method thus requires the nodes to be part of the space, a dated phylogeny (chronogram) and which model to use when slicing through branches rather than tips and nodes.
When a slice occurs not on a tip or a node, six methods are available to select either the descendent or the ancestor's node/tip as an element for this time slice.
For each method, the selected data is: %TODO HL
\begin{itemize}
    \item ``acctran'': always the descendant
    \item ``deltran'': always the ancestor
    \item ``random'': randomly chosen between the descendant or the ancestor
    \item ``proximity'': either the descendant or the ancestor depending on branch length
    \item ``equal.split'': both the descendant and the ancestor with an even probability
    \item ``gradual.split'': both the descendant and the ancestor with a probability depending on branch length
\end{itemize}

Note that the four first models are a proxy for punctuated evolution and the two last ones for gradual evolution.
There is a trade-off between precision and accuracy when using the time-slicing method: a higher number of slices increases the precision of the disparity analysis but also decreases accuracy.
See Guillerme \& Cooper \textit{in review} for full description.

\subsection{Bootstrapping and rarefying}
Disparity measurement can be influenced by sampling \citep{Butler2012}.
To take this source of bias into account, one can bootstrap the multidimensional space or/and to rarefy the data.
Additionally, if disparity is defined as a dimension-level 1 metric, it can be useful to measure it on bootstrapped data to obtain a distribution on which to perform statistical analyses.

Bootstrapping can be achieved by using the \texttt{boot.matrix} function which pseudo-replicates the space following two algorithms (1) the ``full'' algorithm where the bootstrapping is entirely stochastic ($n$ elements are replaced by any $m$ elements drawn from the data) and (2) the ``single'' algorithm where $n = 1$ (somewhat akin to jackknife resampling).

Similarly, rarefaction can be achieved through the same \texttt{boot.matrix} function.
In practice, rarefaction limits the number of elements to be drawn for each bootstrap replication: only $n-x$ elements are selected at each bootstrap replicate (where $x$ is the number of non-sampled elements).

\subsection{Interpreting results}
The functions above all generate a \disp object that can be summarised or plotted using the \texttt{S3} method functions \texttt{summary.dispRity} and \texttt{plot.dispRity}.
These results can also be analysed using the \texttt{test.dispRity} function for comparing subsets or testing hypotheses.

\subsubsection{Summarising and plotting}
The \texttt{summary.dispRity} and \texttt{plot.dispRity} functions allow users to set which central tendency and which quantiles should be represented.
Additionally, the \texttt{plot.dispRity} function graphically represents the summarised results using different representations: (1) ``continuous'' for displaying continuous disparity curves and (2) ``box'', ``lines'', or ``polygons'' to display them using boxplots, confidence interval lines or polygons respectively.
Additional arguments specific to \disp objects can also be used such as \texttt{observed} to display the observed disparity (i.e. non-bootstrapped) or \texttt{rarefaction} to only plot the disparity for a certain number of elements (i.e. the rarefaction level).
The function can also take any additional graphic arguments (\texttt{main}, \texttt{xlab}, \texttt{col}, etc...) from base \texttt{R}.

\subsubsection{Testing hypotheses}
The \texttt{test.dispRity} function allows users to test hypotheses on the disparity data.
Similarly to the \disp function described above, this function can take any test defined by the user or from other R packages.
The \texttt{comparison} arguments indicates in which order (if any) the tests should be applied to the subsets: (1) ``pairwise'' for pairwise comparisons; (2)``referential'' for comparing the first subset to all the others; (3) ``sequential'' for comparing subsets sequentially (e.g. first against second, second against third, etc.); (4) `all'' for comparing all the subsets simultaneously (i.e. \texttt{disparity $\mathtt{\sim}$ subsets}) or (5) any list of pairs (or more) of subsets to compare.

@@

\section{Examples}
Multivariate analysis can be really useful for looking at multiple aspects of organisms diversity together.
For example, one can also look the diversity of morphologies \citep[or disparity;][]{foote1991morphological}.
Using disparity, it is then also possible to assess whether one ecosystem or/and time period display more morphological variation.
The following example is based on a classical morphological disparity analysis.
Note that more examples are available in the package manual (\url{https://rawgit.com/TGuillerme/dispRity/master/inst/gitbook/_book/index.html}).

@@

In this example, the space is defined as a morphospace: the ordination of the distances among discrete morphological characters for 50 mammal species \citep[from][]{beckancient2014}.
Additionally, we can define disparity as the sum of the variances on each dimension \citep{foote1991morphological,Wills1994} that will represent an aspect of the the volume of the morphospace.

@@
